Covering Basis: Comprehensive Model Agnostic Probes of the Binary Black Hole Population

### 2022-10-04 src/tex/results.tex
more comments fixing up text/wording/grammar

We use hierarchical Bayesian inference (see Appendix \ref{sec:hierarchical_inference}) to simultaneously infer the astrophysical mass, spin and redshift distributions of 
a binary black holes (BBHs) given a catalog of gravitational wave observations. We parameterize the binaries' masses with the primary (defined as the more massive component) mass 
and the mass ratio -- defined as $q=\frac{m_2}{m_1}$, bounded in the range from 0 to 1. Furthermore, we model 4 of the 6 total 
spin degrees of freedom of a binary merger with each component spin magnitude $a_1$ and $a_2$, along with the cosine of the tilt angle of each component, 
$\cos{\theta_1}$ and $\cos{\theta_2}$. The tilt angle is defined as the angle between each components' spin vector and the binaries orbital angular momentum vector. 
We assume the polar spin angles are uniformly distributed in the orbital plane. While we choose to use M-Spline distributions to model the primary mass and spin distributions, we 
model the mass ratio distribution with a power law, i.e. $p(q | m_1, m_\mathrm{min}, \beta) \propto q^{\beta} \Theta(qm_1 - m_\mathrm{min}) \Theta(m_1 - qm_1)$. 
Where $\beta$ is the power law index and $\Theta$ is the Heaviside step function which ensures $m_2$ is within the allowed range, [$m_\mathrm{min}$, $m_1$] \citep{Talbot_2018,o1o2_pop,o3a_pop}.
Additionally, we fit a population model on the redshift or luminosity distance distribution of BBHs, assuming a $\Lambda\mathrm{CDM}$ cosmology defined by the parameters 
from the Planck 2015 results \citep{Planck2015}. This defines an analytical mapping between each event's inferred luminosity distance, and it's redshift, which we now use interchangeably. 
We model the evolution of the merger rate with redshift with the \textsc{PowerlawRedshift} model, described by $p(z|\lambda_z)\propto \frac{dV_c}{dz}(1+z)^{\lambda_z-1}$ \citep{Fishbach_2018redshift}. 
We describe our specific prior choices on the hyperparameters and each model in more detail in Appendix \ref{sec:modelpriors}. Following the same cut on the recent GWTC-3 catalog done in the LVK's 
accompanying BBH population analysis, we have 70 possible BBH mergers with false alarm rates less than 1 per year \citep{GWTC3,o3b_astro_dist}. Since it was concluded to be an outlier of the rest of 
the BBH population in both GWTC-2 and GWTC-3, we choose to omit the poorly understood event, GW190814 \citep{190814disc,o3a_pop,o3b_astro_dist,Essick_2022}. This leaves us with 
a catalog of 69 confident BBH mergers observed over a period of about 2 years, from which we want to infer its population properties. Our study provides the first comprehensive 
data-driven investigation of the BBH mass and spin distributions, uncovering new insights and corroborating those found with other methods. 
We start by looking towards our model's inferences of the mass distribution.

@@

We apply our newly constructed models to the BBH analysis that was done in the astrophysical population companion paper to GWTC-3. 
Figure \ref{fig:mass_distribution} shows how the M-Spline model (red) found consistent features in the mass distribution with 
what was inferred with the \textsc{PowerlawPeak} and \textsc{PowerlawSpline} mass models \citep{Talbot_2018,o3a_pop,Edelman_2022ApJ,o3b_astro_dist}. 
In particular our M-Spline model finds a peak in merger rate density as a function of primary mass, at both $\sim10\msun$ and $\sim35\msun$ agreeing with those 
reported using the same dataset in \citet{o3b_astro_dist}. We also find a shallow mass distribution that has considerably larger support for higher mass mergers when 
compared to power-law based models. The M-Spline model finds that 75\% percent of primary masses are below \result{$\CIPlusMinus{\macros[MassDistribution][MSpline][m_75percentile]}$}, 
compared to \result{$\CIPlusMinus{\macros[MassDistribution][PLPeak][m_75percentile]}$} and \result{$\CIPlusMinus{\macros[MassDistribution][PLSpline][m_75percentile]}$}, found with 
the \textsc{PowerlawPeak} and \textsc{PowerlawSpline} models respectively. An interesting consequence of this is that we do not find a confidently evolving merger rate with redshift, as inferred with power-law based mass models in \citet{o3b_astro_dist}. 
There is a strong correlation measured between the primary mass power law index, $\alpha$, and the redshift power-law index, $\lambda_z$. When one infers a steeper mass distribution -- 
the inferred rate must drastically increase with redshift to explain the highest mass events in the catalog. However, when one infers a 
flatter distribution like, as the M-Spline does in figure \ref{fig:mass_distribution}, the higher-mass mergers are consistent with 
the local merger rate density as a function of mass given the selection effects. With our more flexible mass model, we find that the merger rate 
is consistent with being uniform with co-moving volume, with $\lambda_z = $\result{$\CIPlusMinus{\macros[MSplineIIDCompSpins][lamb]}$}, compared with 
\citet{o3b_astro_dist}'s reported value of $\lambda_z = $\result{$\CIPlusMinus{\macros[PLPeak][lamb]}$} with the \textsc{PowerlawPeak} mass model.

@@

The \textsc{Default} spin model describes the spin magnitude of both components are identical and independently distributed (IID) non-singular Beta distributions. \bruce{cite Wysocki/Talbot}
The Beta distribution provides a simple 2-parameter model that can produce a wide range of functional forms on the unit interval. However, the constraint that keeps 
the Beta distribution non-singular (i.e. $\alpha>1$ and $\beta>1$) enforces a spin magnitude that always has $p(a_i=0) = 0$. In reality, it is expected that 
all physical bodies must have at least some non-zero angular momentum, studies have proposed the existence of a distinct subpopulation of non-spinning or 
negligibly spinning black holes that can elude discovery with such a model \citep{FullerMa2019,RouletGWTC2Pop,BuildBetterSpinModels,Callister_NoEvidence,GWTC3MonashSpin}.

>>>>

We use hierarchical Bayesian inference (see Appendix \ref{sec:hierarchical_inference}) to simultaneously infer the astrophysical mass, spin and redshift distributions of 
binary black holes (BBHs) given a catalog of gravitational wave observations. We parameterize the binaries' masses with the primary (defined as the more massive component) mass 
and the mass ratio -- defined as $q=\frac{m_2}{m_1}$, bounded in the range from 0 to 1. Furthermore, we model 4 of the 6 total 
spin degrees of freedom of a binary merger with each component spin magnitude $a_1$ and $a_2$, along with the cosine of the tilt angle of each component, 
$\cos{\theta_1}$ and $\cos{\theta_2}$. The tilt angle is defined as the angle between each components' spin vector and the binaries orbital angular momentum vector. 
We assume the polar spin angles are uniformly distributed in the orbital plane. While we choose to use M-Spline distributions to model the primary mass and spin distributions, we 
model the mass ratio distribution with a power law, i.e. $p(q | m_1, m_\mathrm{min}, \beta) \propto q^{\beta} \Theta(qm_1 - m_\mathrm{min}) \Theta(m_1 - qm_1)$, where $\beta$ is 
the power law index and $\Theta$ is the Heaviside step function which ensures $m_2$ is within the allowed range, [$m_\mathrm{min}$, $m_1$] \citep{Talbot_2018,o1o2_pop,o3a_pop}.
Additionally, we fit a population model on the redshift or luminosity distance distribution of BBHs, assuming a $\Lambda\mathrm{CDM}$ cosmology defined by the parameters 
from the Planck 2015 results \citep{Planck2015}. This defines an analytical mapping between each event's inferred luminosity distance, and it's redshift, which we now use interchangeably. 
We model the evolution of the merger rate with redshift with the \textsc{PowerlawRedshift} model, described by $p(z|\lambda_z)\propto \frac{dV_c}{dz}(1+z)^{\lambda_z-1}$ \citep{Fishbach_2018redshift}. 
We describe our specific prior choices on the hyperparameters and each model in more detail in Appendix \ref{sec:modelpriors}. Following the same cut on the recent GWTC-3 catalog done in the LVK's 
accompanying BBH population analysis, we have 70 possible BBH mergers with false alarm rates less than 1 per year \citep{GWTC3,o3b_astro_dist}. Since it was concluded to be an outlier of the rest of 
the BBH population in both GWTC-2 and GWTC-3, we choose to omit the poorly understood event, GW190814 \citep{190814disc,o3a_pop,o3b_astro_dist,Essick_2022}. This leaves us with 
a catalog of 69 confident BBH mergers observed over a period of about 2 years, from which we want to infer its population properties. Our study provides the first comprehensive 
data-driven investigation of the BBH mass and spin distributions, uncovering new insights and corroborating those found with other methods. 
We start by looking towards our model's inferences of the mass distribution.

@@

We apply our newly constructed models to the BBH analysis that was done in the astrophysical population companion paper to GWTC-3. 
Figure \ref{fig:mass_distribution} shows how the M-Spline model (red) found consistent features in the mass distribution with 
what was inferred with the \textsc{PowerlawPeak} and \textsc{PowerlawSpline} mass models \citep{Talbot_2018,o3a_pop,Edelman_2022ApJ,o3b_astro_dist}. 
In particular our M-Spline model finds a peak in merger rate density as a function of primary mass, at both $\sim10\msun$ and $\sim35\msun$ agreeing with those 
reported using the same dataset in \citet{o3b_astro_dist}. We also find a shallow mass distribution that has considerably larger support for higher mass mergers when 
compared to power-law based models. The M-Spline model finds that 75\% percent of primary masses are below \result{$\CIPlusMinus{\macros[MassDistribution][MSpline][m_75percentile]}$}, 
compared to \result{$\CIPlusMinus{\macros[MassDistribution][PLPeak][m_75percentile]}$} and \result{$\CIPlusMinus{\macros[MassDistribution][PLSpline][m_75percentile]}$}, found with 
the \textsc{PowerlawPeak} and \textsc{PowerlawSpline} models respectively. An interesting consequence of this is that we do not find a confidently evolving merger rate with redshift, 
as inferred with power-law based mass models in \citet{o3b_astro_dist}. There is a strong correlation measured between the primary mass power law index, 
$\alpha$, and the redshift power-law index, $\lambda_z$. When one infers a steeper mass distribution, 
the inferred rate must drastically increase with redshift to explain the highest mass events in the catalog. However, when one infers a 
flatter distribution, as the M-Spline does in figure \ref{fig:mass_distribution}, the higher-mass mergers are consistent with 
the local merger rate density as a function of mass given the selection effects. With our more flexible mass model, we find that the merger rate 
is consistent with being uniform with co-moving volume, with $\lambda_z = $\result{$\CIPlusMinus{\macros[MSplineIIDCompSpins][lamb]}$}, compared with 
\citet{o3b_astro_dist}'s reported value of $\lambda_z = $\result{$\CIPlusMinus{\macros[PLPeak][lamb]}$} with the \textsc{PowerlawPeak} mass model.

@@

The \textsc{Default} spin model describes the spin magnitude of both components as identical and independently distributed (IID) non-singular Beta distributions. \bruce{cite Wysocki/Talbot}
The Beta distribution provides a simple 2-parameter model that can produce a wide range of functional forms on the unit interval. However, the constraint that keeps 
the Beta distribution non-singular (i.e. $\alpha>1$ and $\beta>1$) enforces a spin magnitude that always has $p(a_i=0) = 0$. In reality, it is expected that 
all physical bodies must have at least some non-zero angular momentum, studies have proposed the existence of a distinct subpopulation of non-spinning or 
negligibly spinning black holes that can elude discovery with such a model \citep{FullerMa2019,RouletGWTC2Pop,BuildBetterSpinModels,Callister_NoEvidence,GWTC3MonashSpin}.
